Temporal ordering and registration of images in studies of developmental dynamics

### 2015-02-04 image_analysis_paper_development2/SI_text.tex
after yannis comments; before stas comments

Before applying vector diffusion maps, images must be preprocessed so that the Euclidean distance between the image pixels is informative: we want images who are developmentally similar to have relatively small Euclidean distances, and images who are at disparate developmental time points to have relatively large distances. 
%
We must preprocess images to remove any experimental and imaging artifacts, so that most of the variation is relevant to the developmental dynamics.

For our general purposes, the relevant image operations are 
\begin{description}

\item[Intensity normalization] Contrast-limited adaptive histogram equalization (using the \texttt{adapthisteq} function with an $8 \times 8$ tile grid, and a uniform distribution for the intensities with a clip limit of 0.01) is used to normalize the intensities of signals whose absolute intensity is not meaningful/informative.
%
\item[Blur] The \texttt{imfilter} function with a disc filter is used to blur signals whose small-scale structure is not informative. For example, 
%
\item[Intensity scaling (multichannel images only)] The \texttt{immultiply} function is used to scale signal intensities. This is important for multichannel images, and it determines the (relative) contributions of the various signals.  
%
\item[Mean-centering]  The Canny method \citep{canny1986computational} is used to detect the edges of the object in each image (using the \texttt{edge} function). The image is then translated so that the object (as determined by extremeties of the detected edges) is centered.
%
\item[Size scaling] For images whose relative size is unimportant to the developmental dynamics, the images are rescaled/dialated so that the object size (as determined by extremeties of the detected edges) is constant (we set this to be 80\% of the total image).
%
\item[Removing corners] The image is cropped/filtered using a disc centered in the middle of the image and whose diameter is equal to the number of pixels; any pixels outside of this disc are set to 0 (no intensity). This removes any corner effects under rotations. 
\end{description}

\subsubsection*{{\em Drosophila} live imaging}

%
The original image resolution is $512 \times 512$ for the live {\em Drosophila} images.
%
All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
%
Images were normalized, and then blurred with a filter of radius 5\% of the total image (5 pixels). 
%
Images were not mean-centered or rescaled, as the entire live imaging data set was approximately centered already.

\subsubsection*{Zebrafish images}

The original image resolution was $320 \times 288$ for the zebrafish images.
%
All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
%
Images were not normalized or blurred.
%
Images were mean-centered so that the embryo in each frame was (approximately) centered. 
%
Images were not rescaled for size, as changes in overall were are important.

\subsubsection*{{\em Drosophila} fixed images}

The original image resolution was $1024 \times 1024$ for the fixed {\em Drosophila} images.
%
All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
%
The nuclear channel was normalized, and all channels blurred with a filter of radius 5\% of the total image (5 pixels) to remove the effects of individual nuclei. 
%
The nuclear channel was scaled by half; because this signal occupies a larger fraction of the image relative to the other signals, its overall contribution is large, and so we downscaled it so that each signal would have more comparable weight in the algorithm. 
%
The images were mean-centered and scaled to have a constant size using the nuclear signal to detect the edges of the embryo within the frame.

\subsubsection*{Wing disc images}

The original image resolution was $1024 \times 1024$ for the wing disc z-stacks.
%
Each of the original wing disc z-stacks contains 30--40 images. 
%
All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
%
Each wing disc z-stack was reduced to a 21-image stack, consisting of the brightest image and the 10 images above and below.
%
No channels were normalized or blurred in the images, and the channel intensities were kept at their imaging values. 
%
Images were mean-centered using the Wingless/Patched signal to detect the edges of the sample. 
%
Images were not rescaled for size, as changes in overall were are important.

@@

We demonstrate the algorithms for registration and temporal ordering using a synthetic data set.
%
The relatively simple dynamics of this data set allows us to easily visualize and illustrate the main features of the different algorithms.
%
Motivated by the geometry of our images, we construct a sequence of concentration profiles defined on a ring, and rotate each ring randomly around its center; an example is shown in \fig~\ref{subfig:1d_example}.
%
Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown at the bottom of \fig~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
%
Each concentration profile is a noisy Gaussian (shown in \fig~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as a function of ``time".
%
We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_{100}$, the group of integers modulo $100$).
%
\fig~\ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.

@@

Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional vector space), we want to find a coordinate transformation $y(x)$ that preserves local geometry: points that are ``close" in the original space should also be ``close" in the coordinates $y$.
%
The first step is to construct the matrix $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if points $x_i$ and $x_j$ are ``close.''
%
We use a diffusion kernel,
\begin{equation} \label{eq:dmaps_W}
W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right),
%W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
\end{equation}
where $d(x_i, x_j)$ is a pairwise distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic scale.
%
Points less than $\epsilon$ apart are thus considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
%
$\epsilon$ can be chosen using several techniques (see, for example \citep{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/4-1/2$ of the median of the pairwise distances.

@@

To register sets of images, 
the first step is to compute the optimal alignments between pairs of images.
%
Practically, we have square images discretized as pixels (rather than continuous functions on the plane).
%
For each image pair $I_i$ and $I_j$ we compute
\begin{equation}\label{eq:opt_pairwise}
\theta_{ij} = \argmin_{
0^\circ \le \theta < 360^\circ }
 \|g(I_j, \theta) - I_i \|^2.
\end{equation}
where $g(I_j, \theta)$ is image $I_j$ rotated around the center of the square by $\theta$ degrees.
%
The norm, $\| \cdot \|$, is the Euclidean norm between the pixel intensities of the channels.
%
The domain of the image (a square) is not invariant to our rotations; however, the pixels near the corners of the square are preprocessed to have zero intensity, and so the norm can be meaningfully computed as long as the main image does not ``move out of'' the original square.
%
Image rotation is performed with the \texttt{imwarp} function in Matlab, using linear interpolation to estimate the pixel intensities after rotation.
%
The missing pixels in the corners of the rotated image are taken to have zero intensity.
%
The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
%
Therefore, instead of using an optimization procedure, we discretize the search space and exhaustively search to find the solution (for the results presened, we use $10^\circ$ discretization steps).
%
%We discritize the search space of rotations into 20 possible rotations %($d\theta  \in \{0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10 \}$),
%and 11 possible translations in both . %($dx, dy \in \{-20, -16, -12, \dots, 12, 16, 20 \}$).
%
%We check all possible combinations for the rotation and translations that best align $I_j$ to $I_i$.
%
Although computationally demanding, this ``embarrassingly parallelizable" direct enumeration approach is not prohibitive here.
%
Once we have computed $\theta_{ij}$ for all image pairs, we can proceed with the vector diffusion maps procedure.
%
The rotation matrices returned from vector diffusion maps can be used to calculate the angle of rotation for each image (by inverting \eqref{eq:R_theta}), and the function $g$ as described previously is then used to rotate the images.

@@

%
%\begin{figure*}
%\includegraphics{figS3}
%\caption{Eigenvalue spectra. {\it (A)} Eigenvalue spectrum for the live imaging data set presented in \fig~3. Note that there is a gap after the third eigenvalue. {\it (B), (C)} Second and third embedding coordinates, respectively, versus the first embedding coordinate for the live imaging data set presented in \fig~3. Note that coordinates 2 and 3 are higher harmonics (and thus simple functions) of coordinate 1, and are therefore not informative about structure in the data set.  We can conclude that the data set is effectively one-dimensional. {\it (D)} Eigenvalue spectrum for the live imaging data set presented in \fig~4. Note that there is a gap after the first eigenvalue; we can conclude that the data set is effectively one-dimensional. }
%\end{figure*}
%
%\begin{figure*}
%\includegraphics{figS4}
%\caption{Live imaging data set. {\it (A)} Live imaging data set used for the results in \fig~3. The images have been preprocessed as described above, each image has been arbitrarily rotated, and the order of the images has been shuffled. {\it (B)} Data set from A after registration and ordering. The real time for each image is indicated. }
%\end{figure*}
%
%\begin{figure*}
%\includegraphics{figS5}
%\caption{{\it (A-D)} Four additional live imaging data sets used for the analysis in \fig~3. }
%\end{figure*}
%
%\begin{figure*}	
%\includegraphics{figS6}
%\caption{Fixed embryo imaging data set. {\it (A)} Fixed images used for the results in \fig~4. The images have been preprocessed as described above. {\it (B)} Data set from A after registration and ordering. The rank provided by a trained embryologist for each image is indicated. }
%\end{figure*}

>>>>

Before applying vector diffusion maps, images must be preprocessed so that the Euclidean distance between the image pixels is informative: we need images who are developmentally similar to have relatively small Euclidean distances, and images who are at disparate developmental time points to have relatively large distances. 
%
We must preprocess images to remove any experimental and imaging artifacts, so that most of the variation is relevant to the developmental dynamics.

The relevant image operations are listed below for our general purposes, along with the relevant MATLAB functions. 
\begin{description}

\item[Intensity normalization] Contrast-limited adaptive histogram equalization (using the \texttt{adapthisteq} function with an $8 \times 8$ tile grid, and a uniform distribution for the intensities with a clip limit of 0.01) is used to normalize the intensities of signals whose absolute intensity is not meaningful/informative.
%
\item[Blur] The \texttt{imfilter} function with a disc filter is used to blur signals whose small-scale structure is not informative. 
%
\item[Intensity scaling (multichannel images only)] The \texttt{immultiply} function is used to scale signal intensities. This is important for multichannel images, as it determines the (relative) contributions of the various signals.  
%
\item[Mean-centering]  The Canny method \citep{canny1986computational} is used to detect the edges of the object in each image (using the \texttt{edge} function). The image is then translated so that the object (as determined by extremeties of the detected edges) is centered.
%
\item[Size scaling] For images whose relative size is unimportant to the developmental dynamics, the images are rescaled/dialated so that the object size (as determined by extremeties of the detected edges) is constant (we set this to be 80\% of the total image).
%
\item[Removing corners] The image is cropped/filtered using a disc centered in the middle of the image and whose diameter is equal to the number of pixels; any pixels outside of this disc are set to 0 (no intensity). This removes any corner effects under rotations. 
\end{description}

\subsubsection*{{\em Drosophila} gastrulation (live)}

%
The original image resolution is $512 \times 512$ for the live {\em Drosophila} embryo images.
%
All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
%
Images were normalized, and then blurred with a filter of radius 5\% of the total image (5 pixels). 
%
Images were not mean-centered or rescaled, as the entire live imaging data set was approximately centered already.

\subsubsection*{Zebrafish epiboly}

The original image resolution was $320 \times 288$ for the zebrafish images.
%
A $16$-pixel border was removed from the left and right sides to make the images square, and all images were then subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
%
Images were not normalized or blurred.
%
Images were mean-centered so that the embryo in each frame was (approximately) centered. 
%
Images were not rescaled for size, as changes in overall size are important.

\subsubsection*{{\em Drosophila} gastrulation (fixed)}

The original image resolution was $1024 \times 1024$ for the fixed {\em Drosophila} images.
%
All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
%
The nuclei channel was normalized, and all channels were blurred with a filter of radius 5\% of the total image (5 pixels) to remove the effects of individual nuclei. 
%
The nuclei channel was scaled by half; because this signal occupies a larger fraction of the image relative to the other signals, its overall contribution is large, and so we downscaled it so that each signal would have more comparable weight in the algorithm. 
%
The images were mean-centered and scaled to have a constant size using the nuclei signal to detect the edges of the embryo within the frame.

\subsubsection*{{\em Drosophila} wing discs}

The original image resolution was $1024 \times 1024$ for the wing disc z-stacks.
%
Each of the original wing disc z-stacks contains 30--40 images. 
%
All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
%
Each wing disc z-stack was reduced to a 21-image stack, consisting of the brightest image and the 10 images above and below.
%
No channels were normalized or blurred in the images, and the channel intensities were kept at their imaging values. 
%
Images were mean-centered using the Wingless/Patched signal to detect the edges of the sample. 
%
Images were not rescaled for size, as changes in overall size are important.

@@

We demonstrate the algorithms for registration and temporal ordering using a synthetic data set.
%
The relatively simple dynamics of this data set allows us to easily visualize and illustrate the main features of the different algorithms.
%
Motivated by the geometry of our {\em Drosophila} embryo images, we construct a sequence of concentration profiles defined on a ring, and rotate each ring randomly around its center; an example is shown in \fig~\ref{subfig:1d_example}.
%
Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown at the bottom of \fig~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
%
Each concentration profile is a noisy Gaussian (shown in \fig~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as a function of ``time".
%
We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_{100}$, the group of integers modulo $100$).
%
\fig~\ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.

@@

Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional vector space), we want to find a coordinate transformation $y(x)$ that preserves local geometry: points that are ``close" in the original space should also be ``close" in the coordinates $y$.
%
The first step is to construct the matrix $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if points $x_i$ and $x_j$ are ``close.''
%
We use a diffusion kernel,
\begin{equation} \label{eq:dmaps_W}
W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right),
%W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
\end{equation}
where $d(x_i, x_j)$ is a pairwise distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic scale.
%
Points less than $\epsilon$ apart are thus considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
%
$\epsilon$ can be chosen using several techniques (see, for example \citep{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/4$ of the median of the pairwise distances for the two-dimensional images, and $1/2$ of the median of the pairwise distances for the three-dimensional z-stacks.

@@

To register sets of images, 
the first step is to compute the optimal alignments between pairs of images.
%
Practically, we have square images discretized as pixels (rather than continuous functions on the plane).
%
For each image pair $I_i$ and $I_j$ we compute
\begin{equation}\label{eq:opt_pairwise}
\theta_{ij} = \argmin_{
0^\circ \le \theta < 360^\circ }
 \|g(I_j, \theta) - I_i \|^2.
\end{equation}
where $g(I_j, \theta)$ is image $I_j$ rotated around the center of the square by $\theta$ degrees.
%
The norm, $\| \cdot \|$, is the Euclidean norm between the pixel intensities of the channels.
%
The domain of the image (a square) is not invariant to our rotations; however, the pixels near the corners of the square are preprocessed to have zero intensity, and so the norm can be meaningfully computed as long as the main image does not ``move out of'' the original square.
%
Image rotation is performed with the \texttt{imwarp} function in MATLAB, using linear interpolation to estimate the pixel intensities after rotation.
%
The missing pixels in the corners of the rotated image are taken to have zero intensity.
%
The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
%
Therefore, instead of using an optimization procedure, we discretize the search space and exhaustively search to find the solution (for the results presened, we use $10^\circ$ discretization steps).
%
%We discritize the search space of rotations into 20 possible rotations %($d\theta  \in \{0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10 \}$),
%and 11 possible translations in both . %($dx, dy \in \{-20, -16, -12, \dots, 12, 16, 20 \}$).
%
%We check all possible combinations for the rotation and translations that best align $I_j$ to $I_i$.
%
Although computationally demanding, this ``embarrassingly parallelizable" direct enumeration approach is not prohibitive here.
%
Once we have computed $\theta_{ij}$ for all image pairs, we can proceed with the vector diffusion maps procedure.
%
The rotation matrices returned from vector diffusion maps can be used to calculate the angle of rotation for each image (by inverting \eqref{eq:R_theta}), and the function $g$ as described previously is then used to rotate the images.

@@

\begin{figure}
\textsf{
\begin{tabular}{| r | c | c |}
\hline
Data set & \parbox[c]{4cm}{ \centering VDM Rank Correlation} & \parbox[c]{4cm}{ \centering PCA Rank Correlation} \\
\hline
{\em Drosophila} gastrulation (live) & 0.9989 & 0.8137 \\
Zebrafish epiboly & 0.9955 & 0.6351 \\
{\em Drosophila} gastrulation (fixed) & 0.9716 & 0.8658 \\
{\em Drosophila} wing discs & 0.9436 & 0.9381 \\
\hline
\end{tabular}
}
\caption{Comparison between rank correlation coefficients when ordering using the first vector diffusion maps (VDM) embedding coordinate, and rank correlation coefficients when ordering by the first principal component analysis (PCA) \citep{shlens2005tutorial} projection coefficient. For ordering using PCA, we computed the first principal component of the registered images, and then ordered the images by the projection coefficients onto this first mode. The PCA ordering is always less accurate than the vector diffusion maps ordering. In the zebrafish data set, the ordering is much less accurate using PCA, as the dynamics of the morphing and spreading of the cell mass are highly nonlinear. The PCA and VDM orderings are comparable for the wing disc data; this is to be expected, as the tissue simply grows in time.  }
\end{figure}

### 2015-02-04 image_analysis_paper_development2/code_for_distribution/README/README.tex
after yannis comments; before stas comments

@@

There are four analysis options
\begin{description}
%
\item[Register] (this option is only for two-dimensional images) This registers the imaging data set using vector diffusion maps to compute the optimal rotations. 
%
\item[Order] This orders the imaging data set using diffusion maps. {\em Important:} This option assumes that the imaging data set is preregistered! This option is intended for imaging data sets that were preregistered using some {\em a priori} knowledge about the system. 
%
\item[Register + Order] (this option is only for two-dimensional images) This registers and orders the imaging data set using vector diffusion maps to compute the optimal rotations and the temporal ordering. 
%
\item[Register + Order z-stacks] (this option is only for three-dimensional z-stacks) This registers and orders the imaging data set. It first registers the z-stacks by computing the optimal rotations to align the (two-dimensional) maximum intensity projections using vector diffusion maps.  It then orders the data using diffusion maps on the (now registered) three-dimensional z-stacks. 
%
\end{description}

@@

\subsection{{\em Drosophila} live imaging}

@@

\subsection{Zebrafish}

@@

\subsection{{\em Drosophila} fixed images}

@@

\subsection{Wing disc}

>>>>

\section{License}

This code is covered under the BSD License.\\
Copyright (c) 2015, Carmeline Dsilva\\
All rights reserved.

Redistribution and use in source and binary forms, with or without
modification, are permitted provided that the following conditions are
met:

\begin{enumerate}

    \item Redistributions of source code must retain the above copyright
      notice, this list of conditions and the following disclaimer.
    \item Redistributions in binary form must reproduce the above copyright
      notice, this list of conditions and the following disclaimer in
      the documentation and/or other materials provided with the distribution
    \item Neither the name of the Princeton University nor the names
      of its contributors may be used to endorse or promote products derived
      from this software without specific prior written permission.
\end{enumerate}
THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS ``AS IS''
AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE
ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT OWNER OR CONTRIBUTORS BE
LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR
CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF
SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS
INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN
CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE)
ARISING IN ANY WAY OUT OF THE USE OF THIS SOFTWARE, EVEN IF ADVISED OF THE
POSSIBILITY OF SUCH DAMAGE.

@@

There are four analysis options
\begin{description}
%
\item[Register] (this option is only for two-dimensional images) This registers the imaging data set using vector diffusion maps to compute the optimal rotations. 
%
\item[Order] This orders the imaging data set using diffusion maps. {\em Important:} This option assumes that the imaging data set is preregistered! This option is intended for imaging data sets that were preregistered using some {\em a priori} knowledge about the system. 
%
\item[Register + Order] (this option is only for two-dimensional images) This registers and orders the imaging data set using vector diffusion maps to compute the optimal rotations and the temporal ordering. 
%
\item[Register + Order z-stacks] (this option is only for three-dimensional z-stacks) This registers and orders the imaging data set. It first registers the z-stacks by computing the optimal rotations to align the (two-dimensional) maximum intensity projections along the z-axis using angular synchronization. It then orders the data using diffusion maps on the (now registered) three-dimensional z-stacks. 
%
\end{description}

@@

\subsection{{\em Drosophila} gastrulation (live)}

@@

\subsection{Zebrafish epiboly}

@@

\subsection{{\em Drosophila} gastrulation (fixed)}

@@

\subsection{{\em Drosophila} wing discs}
